Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-06-25 numerical-chemistry.tex
Section 6 edits.

The primordial non-equilibrium chemistry routines used in \enzo\ were first
described by \citet{abel97} and \citet{anninos97}, but have
since been extended with updated reaction rates and the inclusion of
deuterium species \citep{2009PhDT.........5T}.  These routines follow the
non-equilibrium chemistry of a gas of primordial composition with 12 total
species: H, \Hp, He, \Hep, \Hepp, \Hm, \HHp, \HH, e$^-$, D, \Dp, and HD.  \enzo\
also computes the radiative heating and cooling of the gas from atomic and
molecular line excitation, recombination, collisional excitation, free-free
transitions, Compton scattering of the cosmic microwave background, as well as
several models for a metagalactic UV background that heat the gas via
photoionization and photodissociation; see Section~\ref{sec.num.cooling} for
more details.  The chemical and thermal states of the gas can be updated either
at the same hydrodynamical timestep (i.e., de-coupled and operator-split) or
through the same subcycling system (i.e., a coupled chemical and thermal
system).  The default behavior of \enzo\ is to couple these two systems at
subcycles of the hydrodynamic timestep; this results in updates to both the
chemical and thermal states of the gas (which also inform the temperature, the
reaction rate coefficients and the cooling functions of the gas) on timescales
that are faster than those of the gas dynamics.  The gamma used by \enzo\ to
compute the temperature of the gas from the energy and density characteristics
utilizes a variable gamma that includes effects of the rotational state of
molecular hydrogen, enabling it to vary from $5/3$ (fully-atomic) to $7/5$
(fully molecular).  This further coupling of the chemical and thermal states of
the gas underscores the need for coupled chemistry and radiative cooling
solutions.

@@

\enzo\ solves the rate equations using a method based on a semi-implicit
formulation in order to provide a stable, positive definite and first-order
accurate solution.  The update discretization splits chemical changes into
formation components and destruction components and updated with a mixed set of
time states, as described in \citet{anninos97}.  The formation components of
species $S_i$ are computed at the current subcycle time, where the contribution
of species $S_i$ to its own destruction components are computed at the updated
time; all other contributions to the destruction component are computed at the
current time.  This mixed state improves accuracy, ensures species values
are positive definite, and is equivalent to one Jacobi iteration of an implicit
Euler solve.  This technique is optimized by taking the chemical intermediaries
\Hm and \HHp, which have large rate coefficients and low concentrations, and
grouping them into a separate category of equations.  Due to their fast
reactions, these species are very sensitive to small changes in the more
abundant species and are at all times in astrophysical calculations close to
equilibrium values.  Attempting to resolve their formation and destruction
times would necessitate extremely small timesteps, and while these species
provide necessary channels to facilitate the formation of molecular hydrogen at
low densities, they do not otherwise significantly influence the abundance of
species with higher concentrations.  Therefore, reactions governing these two
species can be decoupled from the rest of the network and treated independently
through analytic solutions for equilibrium values.  This allows a large speedup
in solution as the timestepping scheme is applied only to the slower 7- or
10-species network (depending on whether deuterium is included or not), which
will be much closer to the overall hydrodynamic timestep of the simulation.

Even so, the accuracy and stability of the scheme is maintained by subcycling
the rate solver within a single hydrodynamic timestep.  These subcycle
timesteps are determined so that the estimated fractional change in the
electron concentration is limited to no more than $10\%$ per timestep;
additional criteria may be applied based on the expected change from radiative
cooling, chemical heating from the formation of molecular hydrogen.

>>>>

The primordial non-equilibrium chemistry routines used in \enzo\ were first
described by \citet{abel97} and \citet{anninos97}, but have
since been extended with updated reaction rates and the inclusion of
deuterium species \citep{2009PhDT.........5T}.  These routines follow the
non-equilibrium chemistry of a gas of primordial composition with 12 total
species: H, \Hp, He, \Hep, \Hepp, \Hm, \HHp, \HH, e$^-$, D, \Dp, and HD.  \enzo\
also computes the radiative heating and cooling of the gas from atomic and
molecular line excitation, recombination, collisional excitation, free-free
transitions, Compton scattering of the cosmic microwave background, as well as
several models for a metagalactic UV background that heat the gas via
photoionization and photodissociation (see Section~\ref{sec.num.cooling} for
more details).  The chemical and thermal states of the gas can be updated either
at the same hydrodynamical timestep (i.e., decoupled and operator-split) or
through the same subcycling system (i.e., a coupled chemical and thermal
system).  The default behavior of \enzo\ is to couple these two systems at
subcycles of the hydrodynamic timestep; this results in updates to both the
chemical and thermal states of the gas (which also inform the temperature, the
reaction rate coefficients and the cooling functions of the gas) on timescales
that are faster than those of the gas dynamics.  The gamma used by \enzo\ to
compute the temperature of the gas from the energy and density characteristics
utilizes a variable gamma that includes effects of the rotational state of
molecular hydrogen, enabling it to vary from $5/3$ (fully-atomic) to $7/5$
(fully molecular).  This further coupling of the chemical and thermal states of
the gas underscores the need for coupled chemistry and radiative cooling
solutions.

@@

\enzo\ solves the rate equations using a method based on a semi-implicit
formulation in order to provide a stable, positive definite and first-order
accurate solution.  The update discretization splits chemical changes into
formation components and destruction components and updated with a mixed set of
time states, as described in \citet{anninos97}.  The formation components of
species $S_i$ are computed at the current subcycle time, where the contribution
of species $S_i$ to its own destruction components are computed at the updated
time; all other contributions to the destruction component are computed at the
current time.  This mixed state improves accuracy, ensures species values
are positive definite, and is equivalent to one Jacobi iteration of an implicit
Euler solve.  This technique is optimized by taking the chemical intermediaries
\Hm and \HHp, which have large rate coefficients and low concentrations, and
grouping them into a separate category of equations.  Due to their fast
reactions, these species are very sensitive to small changes in the more
abundant species and are at all times in astrophysical calculations close to
their equilibrium values.  Attempting to resolve their formation and destruction
times would necessitate extremely small timesteps, and while these species
provide necessary channels to facilitate the formation of molecular hydrogen at
low densities, they do not otherwise significantly influence the abundance of
species with higher concentrations.  Therefore, reactions governing these two
species can be decoupled from the rest of the network and treated independently
through analytic solutions for equilibrium values.  This allows a large speedup
in solution as the timestepping scheme is applied only to the slower 7- or
10-species network (depending on whether deuterium is included or not), which
will be much closer to the overall hydrodynamic timestep of the simulation.

Even so, the accuracy and stability of the scheme is maintained by subcycling
the rate solver within a single hydrodynamic timestep.  These subcycle
timesteps are determined so that the estimated fractional change in the
electron concentration is limited to no more than $10\%$ per timestep;
additional criteria may be applied based on the expected change in
internal energy from radiative
cooling and from  chemical heating due to the formation of molecular hydrogen.

### 2013-06-25 numerical-cooling.tex
More minor edits
Section 6 edits.

\enzo\ has multiple methods for computing the energy change from radiative
cooling and heating.  All of them assume the gas can be modeled either as
completely optically thin or with a simple, local approximation to optical
thickness.  Below we describe the methods for computing the cooling rates from
metal-free and metal-enriched gas.  Sample cooling curves for each of \enzo's
primary cooling methods are shown in Figure \ref{fig.cooling_rate}.

@@

As discussed in Section~\ref{sec.num.chemistry}, the set of reactions that
characterize a metal-free gas is simple enough to be computed in
non-equilibrium during even very large simulations.  Similarly, the radiative
cooling of metal-free gas is solved by directly computing the cooling
and heating rates from the following individual processes for atomic H
and He: collisional excitation and ionization, recombination,
free-free emission, Compton scattering off of the cosmic microwave
background (CMB), and photo-heating from a UV metagalactic background.
If the \HH~chemistry network is enabled, the following \HH~
cooling processes are also considered: ro-vibrational transitions
\citep{2008MNRAS.388.1627G,1998A&A...335..403G}, heating and cooling from molecular
formation and destruction \citep{2009Sci...325..601T}, and
collision-induced emission \citep{2004MNRAS.348.1019R}.  If Deuterium
chemistry is enabled, then rotational transitions of HD
\citep{1998A&A...335..403G, 1983ApJ...270..578L} are treated as well.  The
radiative cooling calculation is coupled to the update of the chemistry network
such that they both occur within the same subcycling loop.  This is necessary
in regimes where rapid cooling or change in the ionization state occur, as this
will influence the chemical kinetic rate coefficients through both changes in
energy and the equation of state of the gas.  In addition to the subcycle
timestepping contraints mentioned in Section~\ref{sec.num.chemistry}, the
subcycle timestep is also not permitted to exceed 10\% of the cooling time,
$t_{cool} = e/\dot{e}$.  A metagalactic background affects the gas through both
photo-heating and photo-ionization.  These are treated by including
redshift-dependent photo-ionization and photo-heating rate terms in the
chemistry and cooling equations for \ion{H}{1}, \ion{He}{1}, and \ion{He}{2}.
The black curves in Figure \ref{fig.cooling_rate} show cooling rates for a
metal-free gas with number density $n_{\rm H} = 10^{-4}$ cm$^{-3}$ both with
and without a radiation background.  More detail on the specific UV backgrounds
present in \enzo\ is presented below in Section~\ref{sec.num.rad-homogeneous}.

@@

A more sophisticated method makes use of multidimensional cooling and
heating rate tables computed with the photo-ionization code
\texttt{Cloudy} \citep{1998PASP..110..761F}.  This method, detailed in
\citet{2008MNRAS.385.1443S} and \citet{2011ApJ...731....6S}, works by
using \texttt{Cloudy} to compute the cooling and heating rates from
the metal species only.  The primary assumption made is that of
ionization equilibrium.  The tables can vary along up to five dimensions:
density, metallicity, electron fraction,
redshift (for an evolving metagalactic UV background), and
temperature.  Tables can be created for any abundance pattern for
elements up to atomic number 30 (Zn) and for any incident radiation
field.  Cooling from the standard \enzo\ non-equilibrium cooling module is
applied on top of the metal contributions.  The contribution of metals
to the cooling is computed within the same subcycling loop as the
coupled primordial chemistry and cooling solver.  The blue curves in
Figure \ref{fig.cooling_rate} show cooling rates calculated with this
method for a gas with number density $n_{\rm H} = 10^{-4}$ cm$^{-3}$ and
metallicity of 0.5 Z$_{\odot}$.

\begin{figure}
  \begin{center}
    \includegraphics[width=1.0\textwidth]{figures/cooling_rate.eps}
  \end{center}
  \caption{Radiative cooling rates from the various cooling methods
    available in \enzo.  The black curves show cooling rates from a gas
    with primordial composition using the non-equilibrium chemistry
    network.  The blue curves are for a gas with metallicity of 0.5
    $Z_{\odot}$ computed with the Cloudy cooling method.  The solid
    black and blue lines assume collisional processes only while the
    dashed lines include photo-ionization and photo-heating from a UV
    metagalactic background at $z = 0$ with a gas number density of
    10$^{-4}$ cm$^{-3}$.  The rates shown by the
    dashed lines indicate a net heating below $T \sim 10^{4.5}$, where
    the rapid change in rate is evident (the curve is an absolute
    value so it can be shown on a log plot).  The red curve is the
    tabulated cooling function of \citet{SW87}, which assumes a fully
    ionized gas with metallicity of 0.5 $Z_{\odot}$.}
  \label{fig.cooling_rate}
\end{figure}

>>>>

\enzo\ has multiple methods for computing the energy change from radiative
cooling and heating.  All of them assume that the gas can be modeled either as
completely optically thin or with a simple, local approximation to optical
thickness.  In this section, we describe the methods for computing the cooling rates from
metal-free and metal-enriched gas.  Sample cooling curves for each of \enzo's
primary cooling methods are shown in Figure \ref{fig.cooling_rate}.

@@

As discussed in Section~\ref{sec.num.chemistry}, the set of reactions that
characterize a metal-free gas is simple enough to be computed in
non-equilibrium during even very large simulations.  Similarly, the radiative
cooling of metal-free gas is solved by directly computing the cooling
and heating rates from the following individual processes for atomic H
and He: collisional excitation and ionization, recombination,
free-free emission, Compton scattering off of the cosmic microwave
background (CMB), and photo-heating from a metagalactic ultraviolet background.
If the \HH~chemistry network is enabled, the following \HH~cooling 
processes are also considered: ro-vibrational transitions
\citep{2008MNRAS.388.1627G,1998A&A...335..403G}, heating and cooling from molecular
formation and destruction \citep{2009Sci...325..601T}, and
collision-induced emission \citep{2004MNRAS.348.1019R}.  If Deuterium
chemistry is enabled, then rotational transitions of HD
\citep{1998A&A...335..403G, 1983ApJ...270..578L} are treated as well.  The
radiative cooling calculation is coupled to the update of the chemistry network
such that they both occur within the same subcycling loop.  This is necessary
in regimes where rapid cooling or change in the ionization state occur, as this
will influence the chemical kinetic rate coefficients through both changes in
energy and the equation of state of the gas.  In addition to the subcycle
timestepping contraints mentioned in Section~\ref{sec.num.chemistry}, the
subcycle timestep is also not permitted to exceed 10\% of the cooling time,
$t_{cool} = e/\dot{e}$.  A metagalactic background affects the gas through both
photo-heating and photo-ionization.  These are treated by including
redshift-dependent photo-ionization and photo-heating rate terms in the
chemistry and cooling equations for \ion{H}{1}, \ion{He}{1}, and \ion{He}{2}.
The black curves in Figure \ref{fig.cooling_rate} show cooling rates for a
metal-free gas with number density $n_{\rm H} = 10^{-4}$ cm$^{-3}$ both with
and without a radiation background.  More detail on the specific UV backgrounds
present in \enzo\ is presented in Section~\ref{sec.num.rad-homogeneous}.

@@

A more sophisticated method makes use of multidimensional cooling and
heating rate tables computed with the photo-ionization code
\texttt{Cloudy} \citep{1998PASP..110..761F}.  This method, detailed in
\citet{2008MNRAS.385.1443S, 2011ApJ...731....6S}, works by
using \texttt{Cloudy} to compute the cooling and heating rates from
the metal species only.  The primary assumption made is that of
ionization equilibrium.  The tables can vary along up to five dimensions:
density, metallicity, electron fraction,
redshift (for an evolving metagalactic UV background), and
temperature.  Tables can be created for any abundance pattern for
elements up to atomic number 30 (Zn) and for any incident radiation
field.  Cooling from the standard \enzo\ non-equilibrium cooling module is
applied on top of the metal contributions.  The contribution of metals
to the cooling is computed within the same subcycling loop as the
coupled primordial chemistry and cooling solver.  The blue curves in
Figure \ref{fig.cooling_rate} show cooling rates calculated with this
method for a gas with number density $n_{\rm H} = 10^{-4}$ cm$^{-3}$ and
metallicity of 0.5 Z$_{\odot}$.

\begin{figure}
  \begin{center}
    \includegraphics[width=0.8\textwidth]{figures/cooling_rate.eps}
  \end{center}
  \caption{Radiative cooling rates from the various cooling methods
    available in \enzo.  The black curves show cooling rates from a gas
    with primordial composition using the non-equilibrium chemistry
    network.  The blue curves are for a gas with metallicity of 0.5
    $Z_{\odot}$ computed with the Cloudy cooling method.  The solid
    black and blue lines assume collisional processes only while the
    dashed lines include photo-ionization and photo-heating from a UV
    metagalactic background at $z = 0$ with a gas number density of
    10$^{-4}$ cm$^{-3}$.  The rates shown by the
    dashed lines indicate a net heating below $T \sim 10^{4.5}$, where
    the rapid change in rate is evident (the curve is an absolute
    value so it can be shown on a log plot).  The red curve is the
    tabulated cooling function of \citet{SW87}, which assumes a fully
    ionized gas with metallicity of 0.5 $Z_{\odot}$.}
  \label{fig.cooling_rate}
\end{figure}
